---
title: Summer is Coming
subtitle: AI Tools for R, Shiny, and Pharma
author: Joe Cheng
date: 2024-10-29
format:
  revealjs:
    theme: simple
    transition: slide
    slide-number: true
    chalkboard: true
editor:
  render-on-save: true
---

```{r setup,include=FALSE}
options(cli.width = 68)
```

## ⛔️ What I'm NOT here to talk about

- Is AGI coming?
- Can AI replace humans?
- Are we all going to lose our jobs?
- Will AI take over the world and enslave the human race?

## ⛔️ What I'm NOT here to talk about

- ChatGPT and Copilot (mostly)
- Image generation, self-driving cars, and other non-LLM forms of AI

## ✅ What I AM here to talk about

- Interacting with LLMs from code (i.e., Chat APIs)
- How they might be useful despite their dangers and limitations
- How to use them from R
- How to use them from Shiny

## What is an AI chat?

- A back-and-forth conversation between a **user** and an **assistant**, governed by a **system** of rules
- ChatGPT is one example of an AI chat system
- But underneath ChatGPT is a **powerful chat API** that you can program against
- Today I'll show you how easy it is!

## The power of chat APIs

- ChatGPT and Copilot are great, but...
- Coding against chat APIs is WAY more exciting!
- Even with the limitations of today's LLMs, these are tremendously useful tools

## Anatomy of a chat

```{scss}
.message {
  font-size: 0.8em;
  border-radius: 30px;
  padding: 12px 24px;
  margin-bottom: 0.5em;
  --bgcolor: white;
  --fgcolor: black;
  background-color: var(--bgcolor);
  color: var(--fgcolor);
}
.message-left {
  @extend .message;
  margin-right: 30%;
}
.message-right {
  @extend .message;
  margin-left: 30%;
}
.message-system {
  @extend .message-right;
  --fgcolor: lime;
  --bgcolor: black;
}
.message-user {
  @extend .message-right;
  --bgcolor: #ccc;
}
.message-assistant {
  @extend .message-left;
  --bgcolor: #007aff;
  --fgcolor: white;
}
.badge {
  font-family: var(--r-main-font);
  font-size: 1.4rem;
  text-transform: uppercase;
  font-weight: bold;
}

.dummy {
  font-size: 1.4rem;
  text-transform: uppercase;
  background-color: var(--fgcolor);
  color: var(--bgcolor);
  margin-right: auto;
  border-radius: 100px;
  padding: 0.5rem 1rem;
}
```

::: {.message-user}
<div class="badge">🧑 User</div>
When was the R language first released?
:::

::: {.message-assistant}
<div class="badge">🤖 Assistant</div>
The R language was first released in 1995.
:::

## Anatomy of a chat

::: {.message-user}
<div class="badge">🧑 User</div>
When was the R language first released?
:::

::: {.message-assistant}
<div class="badge">🤖 Assistant</div>
The R language was first released in 1995.
:::

::: {.message-user}
<div class="badge">🧑 User</div>
Who created it?
:::

::: {.message-assistant}
<div class="badge">🤖 Assistant</div>
R was created by Ross Ihaka and Robert Gentleman.
:::

## Anatomy of a chat

::: {.message-system}
<div class="badge">⚙️ System</div>
SHOUT your answers
:::

::: {.message-user}
<div class="badge">🧑 User</div>
When was the R language first released?
:::

::: {.message-assistant}
<div class="badge">🤖 Assistant</div>
THE R LANGUAGE WAS FIRST RELEASED IN 1995!
:::

::: {.message-user}
<div class="badge">🧑 User</div>
Who created it?
:::

::: {.message-assistant}
<div class="badge">🤖 Assistant</div>
R WAS CREATED BY ROSS IHAKA AND ROBERT GENTLEMAN!
:::

## Under the hood

Request 1:

```json
{
  model: "gpt-4o",
  messages: [
    {role: "system", content: "SHOUT your answers"},
    {role: "user", content: "When was the R language first released?"}
  ]
}
```

Response:

```json
{
  message: {
    role: "assistant", content: "The R language was first released in 1995."
  }
}
```

## Under the hood

Request 2:

```{.json code-line-numbers="|4-6|7|"}
{
  model: "gpt-4o",
  messages: [
    {role: "system", content: "SHOUT your answers"},
    {role: "user", content: "When was the R language first released?"},
    {role: "assistant", content: "The R language was first released in 1995."}
    {role: "user", content: "Who created it?"}
  ]
}
```

Response:

```json
{
  message: {
    role: "assistant", content: "R was created by Ross Ihaka and Robert Gentleman."
  }
}
```

# Introducing {elmer}

## Introducing {elmer}

- A new package for R for working with chat APIs
- By Hadley Wickham
- Supports OpenAI, Anthropic, Google Gemini, Ollama, Perplexity, Groq, plus any OpenAI-compatible API
- Prioritizes ease of learning and ease of use

## Getting started

```{r}
#| echo: true
#| cache: true

# Assumes $OPENAI_API_KEY is set

chat <- elmer::chat_openai(
  model = "gpt-4o",
  system_prompt = "Be professional and concise."
)
chat$chat("What was the first rotary engined car sold in the US?")
```

## Multi-turn conversations

```{r}
#| echo: true
#| cache: true

# The `chat` object keeps the conversation history
chat$chat("How about in Japan?")
```

# {shinychat} and the Power of Shiny UI + GenAI (10 minutes)

- **Introducing {shinychat}**: Explain how {shinychat} extends {elmer} to allow users to create chatbot interfaces easily within Shiny.
- **Use Cases**: Discuss several possibilities, such as creating conversational dashboards for Pharma or interactive reports that integrate GenAI with Pharma datasets.
- **Live Demo** (optional): A brief walk-through of a simple Shiny chatbot interface using {shinychat}, showing how it can be customized and extended.

# Responsible Use: Reproducibility and Hallucination Mitigation (5-7 minutes)

- **Challenges**: Discuss common issues like hallucinations in AI responses and the importance of reproducibility in research and development, especially for Pharma.
- **Solutions**: Demonstrate how {elmer} and {shinychat} can be configured to improve model transparency and reliability, ensuring outputs are auditable and reproducible.

# Opportunities in Pharma: Compliance and Innovation (10 minutes)

- **Pharma-specific Scenarios**: Explore how these tools can be applied responsibly in Pharma, such as creating AI-powered assistants for clinical trial management or drug discovery.
- **Compliance Considerations**: Address industry regulations (GxP, FDA guidelines, etc.) and how integrating AI solutions must align with these rules. Discuss how {elmer} can support compliance workflows.
- **Interactive Exploration**: End with some thought-provoking questions for the audience about what kinds of applications they see being enabled by these new tools, especially considering the constraints of the Pharma industry.

# Conclusion and Q&A (5-7 minutes)

- **Recap key points**: Highlight the power of integrating GenAI with R and Shiny for building Pharma solutions.
- **Call to Action**: Encourage the audience to experiment with {elmer} and {shinychat}, and consider the opportunities these tools provide for the Pharma industry.
- **Q&A session**: Open the floor for questions from the audience.



## Mitigating factors

- Is there a human in the loop?
- Are humans likely to understand intuitively that they are interacting with a (flawed) AI?
- Is the problem facing the AI narrow enough that we can assume it will succeed?
- Is there a built-in mechanism that can detect when the AI is failing?
- Is the answer/solution more subjective than objective?
- Are the stakes low enough that an incorrect answer/solution is acceptable?