---
title: Summer is Coming
subtitle: AI Tools for R, Shiny, and Pharma
author: Joe Cheng
date: 2024-10-29
format:
  revealjs:
    theme: simple
    transition: slide
    slide-number: true
    chalkboard: true
    incremental: true
editor:
  render-on-save: true
---

```{r setup,include=FALSE}
options(cli.width = 68)
```

## It's easy to be skeptical about GenAI

* Grandiose claims about what LLMs can do
* Even more grandiose claims about what they will become in the near future
* The e/acc movement 🙄

## Meanwhile, in the real world

* ChatGPT tells us plausible falsehoods
* Copilot writes code that often doesn't work
* Facebook is flooded with AI generated nonsense
* Is this even the future we want?

## Why I think "Summer is Coming"

* Judging LLMs by ChatGPT is like judging the iPhone by its phone app
    * Put compute, touchscreen, sensors, and connectivity in billions of peoples' hands, and interesting things happen
* The real power of LLMs is in their APIs!
    * Every programmer now has access to a magic function that approximates human reasoning!

## Agenda

1. Demo
2. How to call LLMs from R
3. Using LLMs responsibly

# Demos

## Demo 1: Sidebot

## Demo 2: Sidebot for Pharma

## RStudio IDE integrations

::: {.nonincremental}
* {pal} by Simon Couch\
[https://simonpcouch.github.io/pal/](https://simonpcouch.github.io/pal/)

* {chattr} by Edgar Ruiz\
[https://mlverse.github.io/chattr/](https://mlverse.github.io/chattr/)
:::

## Other experiments

::: {.nonincremental}
* Turn recipe e-books and web pages into structured (JSON) data
* GitHub issue auto-labeler, based on project-specific criteria
* Automated ALT text for plots (for visually impaired users)
* Code linter and security analyzer, for almost any language
* Mock dataset generator
* Natural language processing: detecting locations in news articles, extracting emotions from reviews
:::

## Negative results

::: {.nonincremental}
* Look at raw data and summarize/interpret it (without offloading to R or Python)
* Generate complicated regular expressions
* Replace technical documentation for Posit's commercial products
:::

# How to use LLMs from R

## Introducing {elmer}

::: {.nonincremental}
[https://hadley.github.io/elmer/](https://hadley.github.io/elmer/)

- A new package for R for working with chat APIs
- **Compatible:** Works with OpenAI, Anthropic, Google, AWS, Ollama, Perplexity, Groq, and more
- **Powerful:** Designed for multi-turn conversations, streaming, async, and tool calling
- **Easy:** Might be the easiest LLM API client in any language
:::

## Getting started

```{r}
#| echo: true
#| cache: true

# Assumes $OPENAI_API_KEY is set

chat <- elmer::chat_openai(
  model = "gpt-4o",
  system_prompt = "Be terse but professional."
)

chat$chat("When was the R language created?")
```

## Multi-turn conversations

```{r}
#| echo: true
#| cache: true

# The `chat` object keeps the conversation history
chat$chat("Who created it?")
```

## Tool Calling

::: {.incremental}

- Allows LLMs to interact with other systems
- Sounds complicated? It isn't!
- Supported by most of the newest LLMs, but not all (notably, not the new OpenAI `o1` models, yet)

:::

## How It Works

::: {.incremental}
- [Not like this](https://sequencediagram.org/index.html?presentationMode=readOnly#initialData=C4S2BsFMAIAkHsDu1j3uaBjAhucIA7Ac2gBEB5AUQGUA5QTAIAVaReAJwGsAoL7TVNtACqAZ0hsuAW2zBxIXABNIIkEQIhM8AiOgAxAIwA2AOwBBaKZEqRwbAWBcADtjahMIZ-ejxHkAokgZAAtxADp2Ih5RcQBaAD5La1t7AC5oAG8AHQJoaDZ0SDSAIgBXMTYigBps3M17P2BigHUgmQByHWAQlkCu8TzVIOBoAiRoGT0-RGwAT2gABRcOAH4q7IBfLkSQGztgeJ8p3pC2cLYiNIBxSmYAegDg8VvMErY2Btuy24AGACZf-QAVlCACsRFouId-McwhF4ttdqkMkVZJJHEU0oZAZVoEU6gowCAtCIMbiRCUCAQZkVNgjkvs4tE2Gksjk8gVitgrDt6Ws2XVZEiigBJYAdaBY6BKIjvZTjAgKaDkykzQAoBNhoAAjQIlUAAMxKGAUswAhEUNkA) - with the assistant executing stuff
- [Yes like this](https://sequencediagram.org/index.html?presentationMode=readOnly#initialData=C4S2BsFMAIAkHsDu1j3uaBjAhucIA7Ac2gBEB5AUQGVpF4AnAawCgWAHbB0TETg4NHjtIBRJGzAAFpAYA6RkRbZMqBtACqAZ1ksAtpNkhcAE0haQRAiEzwCW6ADEAjADYA7AEFonrRa3A2AJs2rIAtAB8vv6BAgBc0ADeADoE0NAM6JAJAEQArjoMOQA0qem2AqLAuQDqUpIA5A7SMOKSMuoMllKCBEjQkk6iiNgAntAAClxMAPwlZSho4FoJANpEkMAA+ph5DAxVW20tDAC6qQC+LNEgAUHAkaEMCQDCngAy79Ab27v7h8cOgAKHIABgATODnABWHIAShYT0iwmGEhOCgYRASAHFKAAVaAAekBskJfwOAkJBUJEKh0LkACstHYWCixGiOhiiI9CglEjlgJA9OwcglXNDitAchUTGAQHYtKKpVo8gQCKMclckVE-LdYtVoAAlfEaQ0AOSSAqFIrFEqlMrlCqVORVao1VxudwEPNkfIWmSguWwuq9wHmaSwdkF8SlAElgE1oOLoGYiAdzAMCCZoK71YAUAmw0AARhI8qAAGZ5DAmMYAQhylyAA)
  - User asks assistant a question; includes metadata for available tools
  - Assistant asks the user to invoke a tool, passing its desired arguments
  - User invokes the tool, and returns the output to the assistant
  - Assistant incorporates the tool's output as additional context for formulating a response
:::

## How It Works

Another way to think of it:

- The client can perform tasks that the assistant can't do
- Tools put control into the hands of the assistant—it decides when to use them, and what arguments to pass in, and what to do with the results
- Having an "intelligent-ish" coordinator of tools is a surprisingly general, powerful capability!

## Tool Calling in {elmer}

[https://hadley.github.io/elmer/articles/tool-calling.html](https://hadley.github.io/elmer/articles/tool-calling.html)

::: {.nonincremental}
1. Write a function intended for the model to call
2. Add function documentation, including each parameter's type and description
3. Call `elmer::create_tool_def(func)` to generate a tool definition
4. Add that definition to the chat using `chat$register_tool`
:::

## Introducing {shinychat}

[https://github.com/jcheng5/shinychat](https://github.com/jcheng5/shinychat)

* Chatbot UI for Shiny for R
* Designed to integrate with {elmer} or bring any other chat client

## Introducing {shinychat}

```{.r code-line-numbers="|5|9|11|12-13|"}
library(shiny)
library(shinychat)

ui <- bslib::page_fluid(
  chat_ui("chat")
)

server <- function(input, output, session) {
  chat <- elmer::chat_openai(system_prompt = "You're a trickster who answers in riddles")
  
  observeEvent(input$chat_user_input, {
    stream <- chat$stream_async(input$chat_user_input)
    chat_append("chat", stream)
  })
}

shinyApp(ui, server)
```

# Using LLMs responsibly

## What are the dangers?

1. Incorrect and unverifiable answers
2. Uneven reasoning, math, and coding capabilities
3. Lack of interpretability
4. Lack of reproducibility
5. Data security/privacy

## Incorrect and unverifiable answers {.smaller}

1. **Just say no** if correctness is paramount.
2. **Keep a human in the loop:** Allow the user to inspect not just the answers, but the method(s) used by the model to get to the answers.
3. **Automatically verify the LLM's answers:** Find e.g. syntax errors in generated code and send it back if necessary.
4. **Use cases with "squishy" answers:** If correctness is subjective, might be a good fit for LLMs.
5. **Set clear expectations with the user** by making it obvious where cool but unreliable technology is being used.

## Uneven reasoning, math, and coding capabilities

1. **Check your own assumptions** empirically by thoroughly testing the use cases you care about. Use the same model/prompt (and if possible, data) you'll use in production.
2. **Outsource math/stats calculations** to tool calls.
3. **Set clear expectations with your users** that the tool is fallible and some perseverance is required.

## Data security/privacy

* Many companies are loathe to send queries (laden with potentially proprietary code and other trade secrets) to OpenAI and Anthropic.

* "Open" models like Llama are safer, but aren't as smart (yet).

* AWS-hosted Anthropic models and Azure-hosed OpenAI models may be helpful.

# Thank you

::: {.noninteractive}
* {elmer}: [https://hadley.github.io/elmer/](https://hadley.github.io/elmer/)
* {shinychat}: [https://github.com/jcheng5/shinychat](https://github.com/jcheng5/shinychat)
:::
